---
title: "Data Visualization - Mini-Project 2"
author: "Paul Dubois `pdubois6743@floridapoly.edu`"
output: html_document
---


```{r}
library(tidyverse)
library(lubridate)
library(plotly)
```

# Read in data

```{r}
path <- "../data/marathon_results_2017.csv"
marathon <- read_csv(path)
colnames(marathon)
```
### Interactive Plot

#### Top 25 runners

```{r}
p1_data <- marathon %>% 
  group_by(Country) %>% 
  summarise(fastest_time = min(`Official Time`),
            fastest_runner = Name[which.min(`Official Time`)]) %>% 
  arrange(fastest_time) %>% 
  head(15)
```

* plot the data

```{r}
top15 <- ggplot(data = p1_data, mapping = aes(x = fastest_time, y = fct_reorder(Country, fastest_time, .desc = TRUE))) +
  geom_col(aes(fill = Country), color = "lightgray", alpha = 0.3) +
  geom_text(aes(label = `fastest_runner`),  nudge_x = -3000, ) +
  scale_x_continuous(limits = c(0, 11000)) +
  labs(
    title = "Top 15 Runners by Country",
    subtitle = "2017 Boston Marathon",
    x = "Time in Seconds",
    y = ""
  ) +
  theme(
    legend.position = "none",
    plot.title = element_text(hjust=0.5, vjust=2, size=14),
    plot.subtitle = element_text(hjust=0.5, vjust=3)
  )

top15
```

```{r}
ggplotly(top15)
```


### Spacial visualization of the number of racers per state in USA

```{r}
unwanted <- c("AA", "AE","AP","MH","GU","PR","VI","DC")
USA_runners <- 
  marathon %>% 
  filter(Country == "USA") %>% 
  filter(!State %in% unwanted) %>% 
  group_by(State) %>% 
  summarise(num_runners = n())%>% 
  rename(state = State)
```

```{r}
USA_runners %>% 
  filter(state == "MA") %>% 
  pull
```

```{r}
USA_runners %>% 
  filter(!state == "MA") %>% 
  summarise(total_other = sum(num_runners))
```


#### Using `usmap` to create spatial visualization

* I attempted to use the USA shapefiles and `sf` package
  * However, I learned a very hard lesson in trying to manually adjust size and shape of the entir plot margins
  * I could not get it zoomed in enough to see details
* Which lead me to use this package

```{r}
library(usmap)
plot_usmap(data = USA_runners, values = "num_runners") +
  scale_fill_continuous(low = "lightblue", high = "black", name = "Number of Runners", label = scales::comma) + 
  labs(title = "US Participation by State", subtitle = "2017 Boston Marathon") +
  theme(
    plot.title = element_text(hjust=0.5, vjust=0, size=14),
    plot.subtitle = element_text(hjust=0.5, vjust=0),
    legend.position = "right"
    )
```


```{r}
us_participation <- plot_usmap(data = USA_runners, values = "num_runners") +
  scale_fill_continuous(low = "lightblue", high = "black", name = "Number of Runners", label = scales::comma) + 
  labs(title = "US Participation by State", subtitle = "2017 Boston Marathon") +
  theme(
    plot.title = element_text(hjust=0.5, vjust=0, size=14),
    plot.subtitle = element_text(hjust=0.5, vjust=0)
    )

ggplotly(us_participation)
```


### Models

#### How much does age affect your official time

* Hypothesize the relationship between `Age` and `Official Time` to be highly positively correlated
* Why?  
  * As we get older, our marathon times would increase
  * However we can see that in the ages from 20-40 there are many runners who had extrememly fast times
* Conclusion is in general, yes there is a positive correlation, it is not highly correlated however as we can tell from our linear model

```{r}
ggplot(data = marathon, aes(x = Age, y = `Official Time`)) + 
  geom_point( alpha = 0.1) +
  geom_smooth(method = "lm") +
  facet_wrap(~`M/F`)
```

#### Let's take a look at a highly correlated relationship
* Hypothesize the faster a runner reaches halfway, it does not necessarily mean they will complete the marathon faster than others 
* Why?
  * It could be that runners are not focusing enough and run faster during the first half and get too tired to keep up the pace
* Conclusion, for the 2017 Boston Marathon, the relationship between the time it took a runner to get to halfway and finish is highly correlated

```{r}
ggplot(data = marathon, aes(x = Half, y = `Official Time`)) + 
  geom_point( alpha = 0.1) +
  geom_smooth(aes(color = `M/F`), method = "lm") +
  facet_wrap(~`M/F`) +
  theme(
    legend.position = "none"
  )
```

#### Let's take a look at an extreme correlation

* Hypothesize the quicker a runner's pace is, the faster they will finish
* Why?
  * The math just seems to add up here :)
* Conclusion, extreme correlation!

```{r}
ggplot(data = marathon, aes(x = Pace, y = `Official Time`)) + 
  geom_point( alpha = 0.1) +
  geom_smooth(aes(color = `M/F`), method = "lm") +
  facet_wrap(~`M/F`) +
  theme(
    legend.position = "none"
  )
```
